Copulas and their potential for ecology

### 2019-12-17 AER_Responses.tex
working on ref comments editorially

Hi Dan,
I have sent your Adv. Ecol. Res. submission to two peers for review. One has already returned their review (below) and I have spoken to the other who assure me it will be submitted shortly and that they are overall positive about the work. Thus I'm happy to accept this paper pending you addressing the referees' comments and thought I would send the first of these over now so that you can get cracking. I'll forward the other referee's comments shortly once I receive them. I hope this is not too onerous a task, and look forward to receiving your revisions. The final deadline for publication is January 8th 2020 - so please could you submit the final revised version then. Please note minor edits can easily be taken care of in the proof stage and it is more important to get a paper to the publishers by the 8th and then edited the proofs than it is to submit a perfect version late. I hope this makes sense. You'll also shortly receive an email from the online submission system letting you know you MS has been returned to you, this open the submission portal for the revision to be submitted. Once again, thanks for your contribution - the volume is shaping up nicely!
Cheers,
Alex
REFEREE 1:
Review of Ghosh et al., "Copulas and their potential for ecology"

%DAN: started here

@@

>>>Thank you for the positive feedback! It was our goal to get readers to see the usefulness of
copulas and consider them for their own research, so we are very happy to hear this!

@@

>>>Thanks for the suggestion, we have made this change.

@@

>>>Thank you for this suggestion. We have considered at various times the possibility
of splitting the manuscript up
and/or removing some of its content. In the end we decided it is best to keep it all together,
for several reasons. First, the parts of the manuscript (theoretical and empirical, the three
research question, the multiple case studies) reinforce each other, providing better evidence
for the points we are making than any of these parts would on its own. Second,
we selected datasets in an attempt to span multiple disciplines of ecology.
Removing some now would create a less representative collection of datasets across the field, and would weaken
our claim that copula structure seems to be a common phenomenon, broadly, across ecology.
Finally, we selected datasets prior to analysis and we present all results. This was done
to be sure to fairly evaluate the hypothesis that non-Gaussian copula structures are
common in ecology. We did *not* carry out additional analyses on other datasets and
only present the results with interesting copula structure. We would prefer not to
modify the manuscript in such a way as to invalidate this strength of our work.

@@

@@

Referee 2

@@

@@

>>>>

REFEREE 1:

Review of Ghosh et al., "Copulas and their potential for ecology"

@@

>>>Thank you for the positive feedback! It was our goal to get readers to see the usefulness of
copulas and consider them for their own research, so we are very happy to hear we have succeeded!

@@

>>>Thanks for the suggestion, we made this change.

@@

>>>Thank you for this suggestion. We have considered at various times the possibility
of splitting the manuscript up
and/or removing some of its content. In the end we decided it is best to keep it all together,
for several reasons. First, the parts of the manuscript (theoretical and empirical, the three
research question, the multiple case studies) reinforce each other, providing better evidence
for the points we are making than any of these parts would on its own. Second,
we selected datasets in an attempt to span multiple disciplines of ecology.
Removing some now would create a less representative collection 
of datasets across the field, and would weaken
our claim that non-Gaussian and informative 
copula structure seems to be a common phenomenon, broadly, across ecology.
Finally, we selected datasets prior to analysis and we present all results. This was done
to be sure to fairly evaluate the hypothesis that non-Gaussian copula structures are
common in ecology. We did *not* carry out additional analyses on other datasets and
then only present the results with interesting copula structure. We would prefer not to
modify the manuscript in such a way as to invalidate this strength of our work.

@@

>>>Tail associations are introduced in the third paragraph of the Introduction, and four 
conceptual examples of the ecological implications/meaning of copula structure and especially 
of tail associations are given in paragraphs 5-8 of the Introduction. These paragraphs 
serve to provide an initial idea of what a left- or right-tail association would mean 
in ecological terms. Because there is a paragraph between paragraphs 3 and 5, we added 
a note to paragraph 3 indicating that "We give four examples below which illustrate some of the 
potential ecological meaning of tail associations."

@@

REFEREE 2

@@

>>><Shyamolina is building a preliminary version of the table>

@@

>>><Shyamolina is collecting the formulas>

### 2019-12-17 Paper.Rmd
working on ref comments editorially

A main benefit of a copula approach is 
that it can detect associations in the tails of distributions, and asymmetries of such associations. 
Tail association (introduced formally in the "Background on copulas" section, below, section 
\ref{Background}) 
is association between extreme values of variables. 
If smaller values of two positively associated variables are more 
strongly associated than are larger values, the variables are said to
have stronger lower- or left-tail association than right- or upper-tail association
(Fig. \ref{fig_cop_pedag1}C);
and *vice versa* if larger values are more strongly associated than smaller
ones (Fig. \ref{fig_cop_pedag1}D). Datasets of the same 
Spearman or Kendall correlation can have a range of tail associations 
(Fig. \ref{SM-fig_cop_pedag}), which can be quantified (see sections on Background and
methods for Q1 below, sections \ref{Background} and \ref{Methods}). 
Although copulas can be used to model the entire
complex dependence structure of data, we have found tail associations to
usefully describe the full structure of dependence, so tail association 
is one of the main foci of this paper.

@@

One way we may expect, *a priori*, a copula 
study of complex dependence structures between
ecological variables
to better reflect ecological relationships relates to Liebig's law of the
minimum. Liebig's law is 
the idea that growth is controlled not by total resources but by the resources which
are scarcest relative to organism needs. 
If, for instance, the growth of a plant depends on soil 
nitrogen, N, and other factors, a plot of plant growth rates versus soil N may 
look more like Fig. \ref{fig_cop_pedag1}A or C than like 
Fig. \ref{fig_cop_pedag1}B or D. Nitrogen 
will control plant growth, producing a clear
relationship, only when it is limiting, in the left portions of the distributions. 
The portions of the distributions that dominate the association are visible in a rank-based
plot, but may not be revealed by standard correlation techniques.

A second reason why we may expect understanding of complex dependence structures
to benefit ecological
research is that prior work has demonstrated, using copulas, complex 
stucture in the spatial dependence of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018;
@li2013]. One may expect an environmental variable 
measured through time in two locations to show strong tail associations between
the locations if
intense meteorological events are also widespread in their effects, as seems 
frequently to be the case: extreme values are associated with intense events,
so happen in both places at the same time, whereas moderate values of the 
environmental variable may be associated with local phenomena, which differ
between the locations.
Spatial dependence in an environmental variable 
tends to beget spatial dependence 
in fluctuations of populations or other ecosystem variables 
(this is called \emph{spatial synchrony}) which are influenced by the environmental 
variable. This influence is called the \emph{Moran effect}. 
If complex dependence structure is transmitted, 
possibly in modified form, 
from environmental to ecological variables, then we may expect complex dependence 
structure to be a common feature of the 
spatial synchrony of population, community, 
biogeochemical and other environmentally influenced ecological variables.
Synchrony is a phenomenon of substantial interest in 
ecology [@Liebhold2004; @Sheppard2016; @walteretal2017].

A third reason why we may expect an understanding of complex 
dependence structures to benefit 
ecological research is that such an approach may help illuminate 
the causal mechanisms between variables. For instance, if two species
Sp1 and Sp2 are strong competitors, abundances of the two species
across quadrats should be negatively related, as in 
Fig. \ref{fig_cop_pedag1}E-F. If Sp1 is plotted on the horizontal axis
and Sp2 on the vertical, then Fig. \ref{fig_cop_pedag1}E 
may suggest Sp1 is the dominant 
competitor: when Sp1 is abundant, Sp2 is necessarily rare because it is suppressed; 
whereas when Sp1 is rare, Sp2
may be abundant, or may also be relatively rare due to 
limiting factors other than Sp1. 
Alternatively, Fig. \ref{fig_cop_pedag1}F may suggest that Sp2
is the dominant competitor. Other causal hypotheses may produce 
similar dependence structure, and it is usually impossible to obtain
complete information on causal pathways from analyses which are
fundamentally based on associations. However, more detailed 
examination of the patterns of dependence of variables may
well be combined with biological and other reasoning to rule out
some causal hypotheses which could not be eliminated solely 
with standard correlation approaches. @Popovic2018 has previously
used copulas to help illuminate causal aspects of relationships
between species.

A fourth reason why we may expect a copula understanding of complex 
dependence structures to be useful
for ecology has to do with spatially aggregated or averaged quantities. 
Many ecological variables of applied importance 
depend on the spatial average or sum of local 
quantities. For instance, regional methane and CO$_2$ 
fluxes are the sum of local fluxes; 
and the total economic value of a fishery is the sum of local catches. 
We will explore how details, which can be captured using copulas, 
of the dependence between fluctuations of local
quantities can influence fluctuations of 
the spatial mean or sum, and how this may influence 
higher organizational levels in ecology and human concerns.
To illustrate this idea we cite @li2013, who have demonstrated that the overall 
reliability of wind-generated electricity depends sensitively on details
of the dependence between wind speeds at multiple generator sites. 
Spatially aggregated ecological variables may be subject to a similar 
effect, with
possible consequences. For instance, if populations of a 
pest species in different locations are all positively associated and are also more strongly
related to each other in their right tails, then local outbreaks will tend 
to occur together, creating regional epidemics. 
On the other hand, stronger left-tail associations in a pest species, even if overall 
correlation were unchanged, would have more benign effects.

@@

We introduce some standard copula families, as examples of copulas and also
because we will fit these families to data later. The bivariate normal
copula family, which consists of copulas of bivariate normal
distributions, is a one-parameter family. The parameter, which we here call $p$,
corresponds to the correlation
of the related bivariate normal distribution, and controls the degree of association
between the variables. A normal copula with $p=0.7$ was already introduced (Fig.
\ref{PedagogFigCopThry}A). The formula for the
copula is $\Phi_{2,p}(\Phi_{1}^{-1}(u),\Phi_{1}^{-1}(v))$, where $\Phi_1^{-1}$ is the inverse of the
cummulative dstribution function (cdf) of a (univariate) standard 
normal and $\Phi_{2,p}$ is the joint cdf of a
bivariate normal distribution with mean $(0,0)$ and covariance matrix having $1$s on the diagonal
and $p$ in the off diagonal positions. However, formulas for copulas are often not as 
directly informative about copula properties as probability density function (pdf) 
pictures; Fig. \ref{figped_N} has 
example pdfs corresponding to bivariate normal copulas for various values of $p$.
Note that the pdfs are symmetric
around the diagonal line $v=-u+1$ (Fig. \ref{figped_N}), so normal copulas have symmetric 
associations between the two variables in the left and right tails. 
The Clayton copula family,
of which we have already pictured an example (Fig.
\ref{PedagogFigCopThry}B), is another one-parameter family. In contrast to normal copulas, Clayton copulas
have stronger left- than right-tail association. The formula is 
<!--***DAN: is it clear implicitly from what came before that this is the cdf or what?-->
$\left[  \reumax(u^{-p}+v^{-p}-1,0)\right]^{-1/p}$, for parameter $p$; though this again provides 
less direct intuition than do example pdfs (Fig. \ref{figped_C}). Higher values of the parameter,
$p$ produce copulas from the Clayton family with higher Kendall or Spearman correlation. The survival
Clayton family is the symmetric opposite of the Clayton family, showing
stronger right- than left tail association (Fig. \ref{SM-figped_SC}). The BB1 copula 
family is a two-parameter family, thereby providing more flexibility: for some parameters
BB1 copulas have stronger left- than right-tail association, and for others the reverse
(Fig. \ref{figped_BB1}). See @nelsen2006_copula and @joe2014_dependence for further
information on these and other copula families.

>>>>

A main benefit of a copula approach is 
that it can detect associations in the tails of distributions, and asymmetries of such associations. 
Tail association (introduced formally in the "Background on copulas" section, below, section 
\ref{Background}) 
is association between extreme values of variables. 
If smaller values of two positively associated variables are more 
strongly associated than are larger values, the variables are said to
have stronger lower- or left-tail association than right- or upper-tail association
(Fig. \ref{fig_cop_pedag1}C);
and *vice versa* if larger values are more strongly associated than smaller
ones (Fig. \ref{fig_cop_pedag1}D). Datasets of the same 
Spearman or Kendall correlation can have a range of tail associations 
(Fig. \ref{SM-fig_cop_pedag}), which can be quantified (see sections on Background and
methods for Q1 below, sections \ref{Background} and \ref{Methods}). 
Although copulas can be used to model the entire
complex dependence structure of data, we have found tail associations to
usefully describe the full structure of dependence, so tail association 
is one of the main foci of this paper. We give four examples below which
illustrate some of the potential ecological meaning of tail associations.

@@

One way we may expect, *a priori*, a study of tail associations and of the 
complex copula dependence structures between
ecological variables
to better reflect ecological relationships relates to Liebig's law of the
minimum. Liebig's law is 
the idea that growth is controlled not by total resources but by the resources which
are scarcest relative to organism needs. 
If, for instance, the growth of a plant depends on soil 
nitrogen, N, and other factors, a plot of plant growth rates versus soil N may 
look more like Fig. \ref{fig_cop_pedag1}A or C than like 
Fig. \ref{fig_cop_pedag1}B or D, i.e., the two variables may show 
left-tail association. Nitrogen 
will control plant growth, producing a clear
relationship, only when it is limiting, in the left portions of the distributions. 
The portions of the distributions that dominate the association are visible in a rank-based
plot, but may not be revealed by standard correlation techniques.

A second reason why we may expect understanding of complex dependence structures
to benefit ecological
research is that prior work has demonstrated, using copulas, complex 
stucture in the spatial dependence of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018;
@li2013]. One may expect an environmental variable 
measured through time in two locations to show strong tail associations between
the locations if
intense meteorological events are also widespread in their effects, as seems 
frequently to be the case: extreme values are associated with intense events,
so happen in both places at the same time, whereas moderate values of the 
environmental variable may be associated with local phenomena, which differ
between the locations.
Spatial dependence in an environmental variable 
tends to beget spatial dependence 
in fluctuations of populations or other ecosystem variables 
(this is called \emph{spatial synchrony}) which are influenced by the environmental 
variable. This influence is called the \emph{Moran effect}. 
If asymmetric tail associations, or other complex dependence structure, is transmitted, 
possibly in modified form, 
from environmental to ecological variables, then we may expect complex dependence 
structure and tail associations to be a common feature of the 
spatial synchrony of population, community, 
biogeochemical and other environmentally influenced ecological variables.
Synchrony is a phenomenon of substantial interest in 
ecology [@Liebhold2004; @Sheppard2016; @walteretal2017].

A third reason why we may expect an understanding of complex 
dependence structures and tail associations to benefit 
ecological research is that such an approach may help illuminate 
the causal mechanisms between variables. For instance, if two species
Sp1 and Sp2 are strong competitors, abundances of the two species
across quadrats should be negatively related, as in 
Fig. \ref{fig_cop_pedag1}E-F. If Sp1 is plotted on the horizontal axis
and Sp2 on the vertical, then Fig. \ref{fig_cop_pedag1}E 
may suggest Sp1 is the dominant 
competitor: when Sp1 is abundant, Sp2 is necessarily rare because it is suppressed; 
whereas when Sp1 is rare, Sp2
may be abundant, or may also be relatively rare due to 
limiting factors other than Sp1. 
Alternatively, Fig. \ref{fig_cop_pedag1}F may suggest that Sp2
is the dominant competitor. Other causal hypotheses may produce 
similar dependence structure, and it is usually impossible to obtain
complete information on causal pathways from analyses which are
fundamentally based on associations. However, more detailed 
examination of the tail associations and potentially other 
patterns of dependence of variables may
well be combined with biological and other reasoning to rule out
some causal hypotheses which could not be eliminated solely 
with standard correlation approaches. @Popovic2018 has previously
used copulas to help illuminate causal aspects of relationships
between species.

A fourth reason why we may expect an understanding of tail associations and of complex 
dependence structures to be useful
for ecology has to do with spatially aggregated or averaged quantities. 
Many ecological variables of applied importance 
depend on the spatial average or sum of local 
quantities. For instance, regional methane and CO$_2$ 
fluxes are the sum of local fluxes; 
and the total economic value of a fishery is the sum of local catches. 
We will explore how details, which can be captured using copulas and which
relate to tail associations, 
of the dependence between fluctuations of local
quantities can influence fluctuations of 
the spatial mean or sum, and how this may influence 
higher organizational levels in ecology and human concerns.
To illustrate this idea we cite @li2013, who have demonstrated that the overall 
reliability of wind-generated electricity depends sensitively on details
of the dependence between wind speeds at multiple generator sites. 
Spatially aggregated ecological variables may be subject to a similar 
effect, with
possible consequences. For instance, if populations of a 
pest species in different locations are all positively associated and are also more strongly
related to each other in their right tails, then local outbreaks will tend 
to occur together, creating regional epidemics. 
On the other hand, stronger left-tail associations in a pest species, even if overall 
correlation were unchanged, would have more benign effects.

@@

We introduce some standard copula families, as examples of copulas and also
because we will fit these families to data later. The bivariate normal
copula family, which consists of copulas of bivariate normal
distributions, is a one-parameter family. The parameter, which we here call $p$,
corresponds to the correlation
of the related bivariate normal distribution, and controls the degree of association
between the variables. A normal copula with $p=0.7$ was already introduced (Fig.
\ref{PedagogFigCopThry}A). The formula for the
copula is $\Phi_{2,p}(\Phi_{1}^{-1}(u),\Phi_{1}^{-1}(v))$, where $\Phi_1^{-1}$ is the inverse of the
cummulative dstribution function (cdf) of a (univariate) standard 
normal and $\Phi_{2,p}$ is the joint cdf of a
bivariate normal distribution with mean $(0,0)$ and covariance matrix having $1$s on the diagonal
and $p$ in the off diagonal positions. However, formulas for copulas are often not as 
directly informative about copula properties as probability density function (pdf) 
pictures; Fig. \ref{figped_N} has 
example pdfs corresponding to bivariate normal copulas for various values of $p$.
Note that the pdfs are symmetric
around the diagonal line $v=-u+1$ (Fig. \ref{figped_N}), so normal copulas have symmetric 
associations between the two variables in the left and right tails. 
The Clayton copula family,
of which we have already pictured an example (Fig.
\ref{PedagogFigCopThry}B), is another one-parameter family. In contrast to normal copulas, Clayton copulas
have stronger left- than right-tail association. The formula is 
$\left[  \reumax(u^{-p}+v^{-p}-1,0)\right]^{-1/p}$, for parameter $p$; though this again provides 
less direct intuition than do example pdfs (Fig. \ref{figped_C}). Higher values of the parameter,
$p$ produce copulas from the Clayton family with higher Kendall or Spearman correlation. The survival
Clayton family is the symmetric opposite of the Clayton family, showing
stronger right- than left tail association (Fig. \ref{SM-figped_SC}). The BB1 copula 
family is a two-parameter family, thereby providing more flexibility: for some parameters
BB1 copulas have stronger left- than right-tail association, and for others the reverse
(Fig. \ref{figped_BB1}). See @nelsen2006_copula and @joe2014_dependence for further
information on these and other copula families.
